\begin{section}{Introduction to Quantification}

The sentence ``$x>0$'' is not itself a proposition because $x$ is a \textbf{free variable}. A sentence with a free variable is a \textbf{predicate}. To turn a predicate into a proposition, we must either substitute values for each free variable, or else ``quantify'' the free variables.

%\begin{exercise}
%Give 3 examples of mathematical predicates involving 1, 2, and 3 free variables, respectively.
%\end{exercise}

Function notation is a convenient way to represent predicates. For example, each of the following represents a predicate with the indicated free variables.
\begin{itemize}
\item $S(x):=``x^2-4=0"$
\item $L(a,b):=``a<b"$
\item $F(x,y):=``x \mbox{ is friends with } y"$
\end{itemize}
The notation $:=$ indicates a definition.  Also, note that the use of the quotation marks above removed some ambiguity.  What would $S(x)=x^2-4=0$ mean?  It looks like $S(x)$ equals 0, but actually we want $S(x)$ to represent the whole sentence ``$x^2-4=0$". 

One way we can make propositions out of predicates is by assigning specific values to the free variables.  That is, if $P(x)$ is a predicate and $x_0$ is specific value for $x$, then $P(x_0)$ is now a proposition (and is either true or false).

\begin{exercise}
Consider $S(x)$ and $L(a,b)$ as defined above. Determine the truth values of $S(0)$, $S(-2)$, $L(2,1)$, and $L(-3,-2)$. Is $L(2,b)$ a proposition or a predicate?  Explain.
\end{exercise}

Besides substituting specific values for free variables in a predicate, we can also make a claim about which values of the free variables apply to the predicate.

\begin{exercise}\label{ex:quantified predicates}
Both of following sentences are propositions. Decide whether each is true or false. What would it take to justify your answers?
\begin{enumerate}[label=\textrm{(\alph*)}]
\item For all $x\in\mathbb{R}$, $x^2-4=0$.\footnote{The symbol $\mathbb{R}$ denotes the set of all real numbers.}
\item There exists $x\in\mathbb{R}$ such that $x^2-4=0$.
\end{enumerate}
\end{exercise}

\begin{definition}
``For all" is the \textbf{universal quantifier} and ``there exists\ldots such that" is the \textbf{existential quantifier}.
\end{definition}

We can replace ``there exists\ldots such that" with phrases  like ``for some" (possibly with some other tweaking to the sentence). Similarly, ``for all", ``for any", ``for every" are used interchangeably in mathematics (even though they might convey slightly different meanings in colloquial language). It is important to note that the existential quantifier is making a claim about ``at least one" \emph{not} ``exactly one." 

Variables that are quantified with a universal or existential quantifier are said to be \textbf{bound}.  To be a proposition, \emph{all} variables must be bound. That is, in a proposition all variables are quantified. 

We must take care to specify the universe of acceptable values for the free variables. Consider the sentence ``For all $x$, $x>0$.'' Is this sentence true or false? The answer depends on what set the universal quantifier applies to. Certainly, the sentence is false if we apply it for all $x\in\mathbb{Z}$. However, the sentence is true for all $x\in\mathbb{N}$. Context may resolve ambiguities, but otherwise, we must write clearly: ``For all $x\in\mathbb{Z}$, $x>0$'' or ``For all $x\in\mathbb{N}$, $x>0$.'' The set of acceptable values for a variable is the called the \textbf{universe (of discourse)}.

\begin{exercise}
Suppose our universe of discourse is the set of integers.
\begin{enumerate}[label=\textrm{(\alph*)}]
\item Provide an example of a predicate $P(x)$ such that ``For all $x$, $P(x)$" is true.
\item Provide an example of a predicate $Q(x)$ such that ``For all $x$, $Q(x)$" is false, but ``There exists $x$ such that $Q(x)$" is true.
\end{enumerate}
\end{exercise}

If a predicate has more than one free variable, then we can build propositions by quantifying each variable.  However, \emph{the order of the quantifiers is extremely important!}

\begin{exercise}\label{exer:ways to quantify}
Let $P(x,y)$ be a predicate with free variables $x$ and $y$ in a universe of discourse $U$. One way to quantify the variables is ``For all $x \in U$, there exists $y \in U$ such that $P(x,y)$.'' How else can the variables be quantified?
\end{exercise}

The possibilities listed in the previous exercise are \emph{not} equivalent to each other.

\begin{exercise}
Suppose the universe of discourse is the set of married people.  Consider the predicate $M(x,y):=``x\mbox{ is married to }y"$.  Discuss the meaning of each of the following.
\begin{enumerate}[label=\textrm{(\alph*)}]
\item For all $x$, there exists $yM$ such that $M(x,y)$.
\item There exists $y$ such that for all $x$, $M(x,y)$.
\item For all $x$, for all $y$, $M(x,y)$.
\item There exists $x$ such that there exists $y$ such that $M(x,y)$.
\end{enumerate}
\end{exercise}

\begin{exercise}
Consider the predicate $F(x,y):=``x=y^2"$.  Discuss the meaning of each of the following.
\begin{enumerate}[label=\textrm{(\alph*)}]
\item There exists $x$ such that there exists $y$ such that $F(x,y)$.
\item There exists $y$ such that there exists $x$ such that $F(x,y)$.
\item For all $x\in \mathbb{R}$, for all $y\in\mathbb{R}$, $F(x,y)$.
\item For all $y\in \mathbb{R}$, for all $x\in\mathbb{R}$, $F(x,y)$.
\end{enumerate}
\end{exercise}

%\begin{exercise}
%Suppose that the universe of discourse is the set of real numbers.  Consider the predicate $G(x,y):=``x>y"$.  Find all possible \emph{distinct} ways to quantify the variables to create propositions and then determine the truth value of each (you do not need to prove your answers).
%\end{exercise}

There are a couple of key points to keep in mind about quantification. To be a proposition, all variables must be quantified.  This can happen in at least two ways:
\begin{itemize}
\item The variables are explicitly bound by quantifiers in the same sentence.
\item The variables are implicitly bound by preceding sentences or by context. Statements of the form ``Let $x=\ldots$" and ``Let $x\in\ldots$" bind the variable $x$ and remove ambiguity.
\end{itemize}
The order of the quantification is important.  Reversing the order of the quantifiers can substantially change the meaning of a proposition.

Quantification and logical connectives (``and,'' ``or,'' ``If\ldots, then\ldots,'' and ``not'') enable complex mathematical statements. For example, the formal definition of $\lim_{x\to c}f(x)=L$ is 
\begin{quote}
For all $\epsilon >0$, there exists $\delta >0$ such that for all $x$, if $0<|x-c|<\delta$, then $|f(x)-L|<\epsilon$.
\end{quote}

In order to study the abstract nature of complicated mathematical statements, it is useful to adopt some notation.

\begin{definition}
The universal quantifier ``for all'' is denoted $\boxed{\forall}$, and the existential quantifier  ``there exists\ldots such that'' is denoted $\boxed{\exists}$.
\end{definition}

Using our abbreviations for the logical connectives and quantifiers, we can symbolically represent mathematical propositions. For example, the (true) proposition ``There exists $x \in \mathbb{R}$ such that $x^2-1=0$'' becomes ``$(\exists x \in \mathbb{R})(x^2-1=0)$,'' while the (false) proposition ``For all $x\in \mathbb{N}$, there exists $y\in\mathbb{N}$ such that $y<x$'' becomes ``$(\forall x)(x\in\mathbb{N}\implies (\exists y)(y\in\mathbb{N}\implies y<x))$'' or ``$(\forall x\in\mathbb{N})(\exists y\in\mathbb{N})(y<x)$.''

\begin{exercise} Convert the following statements into statements using only logical symbols.  Assume that the universe of discourse is the set of real numbers.
\begin{enumerate}[label=\textrm{(\alph*)}]
\item There exists a number $x$ such that $x^2+1$ is greater than zero.
\item There exists a natural number $n$ such that $n^2=36$. 
\item For every real number $x$, $x^2$ is greater than or equal to zero.
\end{enumerate}
\end{exercise}

\begin{exercise}
Express the formal definition of a limit in logical symbols.
\end{exercise}

If $A(x)$ and $B(x)$ are predicates, then it is standard practice for the sentence $A(x)\implies B(x)$ to mean $(\forall x)(A(x)\implies B(x))$ (where the universe of discourse for $x$ needs to be made clear).  In this case, we say that the universal quantifier is implicit.

\begin{exercise}
Consider the proposition ``If $\epsilon >0$, then there exists $N\in\mathbb{N}$ such that $1/N<\epsilon$."  Assume the universe of discourse is the set $\mathbb{R}$.
\begin{enumerate}[label=\textrm{(\alph*)}]
\item Express the statement logical symbols.   Is the statement true?
\item Reverse the order of the quantifiers to get a new statement. Does the meaning change?  If so, how?  Is the new statement true?
\end{enumerate}
\end{exercise}

The symbolic expression $(\forall x)(\forall y)$ can be replaced with the simpler expression $(\forall x,y)$ as long as $x$ and $y$ are elements of the same universe.

\begin{exercise}
Express the statement ``For all $x,y\in\mathbb{R}$ with $x<y$, there exists $m\in\mathbb{R}$ such $x<m<y$" using logical symbols.
\end{exercise}

\begin{exercise}
Rewrite the following statements in words and determine whether each is true or false.
\begin{enumerate}[label=\textrm{(\alph*)}]
\item $(\forall n \in \mathbb{N})(n^2 \geq 5)$
\item $(\exists n \in \mathbb{N})(n^2-1=0)$
\item $(\exists N \in \mathbb{N})(\forall  n > N)(\frac{1}{n} < 0.01)$
\item $(\forall m, n \in \mathbb{Z})(2|m \wedge 2|n \implies 2|(m+n))$
\item $(\forall x \in \mathbb{N})(\exists y \in \mathbb{N})(x-2y=0)$
\item $(\exists x \in \mathbb{N})(\forall y \in \mathbb{N})(y \leq x)$
\end{enumerate}
\end{exercise}

To whet your appetite for the next section, consider how you might prove a statement of the form ``For all $x$\ldots.'' If a statement is false, then its negation is true. How would you go about negating a statement involving quantifiers? 

\end{section}