\begin{section}{Introduction to Functions}

In this section, we will introduce the concept of function as a special type of relation.  Our definition should agree with any previous definition of function that you may have learned.

Up until this point, you may have only encountered functions as an algebraic rule, e.g., $f(x)=x^{2}-1$, for transforming one real number into another.  However, we can study functions in a much broader context.  The basic building blocks of a function are a first set and a second set, say $X$ and $Y$, and a ``correspondence'' that assigns each element of $X$ to exactly one element of $Y$.  Let's take a look at the actual definition.

\begin{definition}\label{def:function}
Let $X$ and $Y$ be two nonempty sets.  A \textbf{function $f$ from $X$ to $Y$} is a relation from $X$ to $Y$ such that for every $x\in X$, there exists a unique $y\in Y$ such that $(x,y)\in f$. The set $X$ is called the \textbf{domain} of $f$ and is denoted by $\dom(f)$.  The set $Y$ is called the \textbf{codomain} of $f$ and is denoted by $\codom(f)$ while the subset of the codomain defined via
\[
\range(f):=\{y\in Y\mid \mbox{there exists }x\mbox{ such that } (x,y)\in f\}
\]
is called the \textbf{range} of $f$ or the \textbf{image of $X$} under $f$.
\end{definition}

We will write $f:X\to Y$ to indicate that $f$ is a function from $X$ to $Y$. We will make use of statements such as ``Let $f:X\to Y$ be the function defined via\ldots" or ``Define $f:X\to Y$ via\ldots", where $f$ is understood to be a function in the second statement. Sometimes the word \textbf{mapping} (or \textbf{map}) is used in place of the word function.  If $(a,b)\in f$ for a function $f$, we often write $f(a)=b$ and say that ``$f$ maps $a$ to $b$'', ``$b$ is the image of $a$ under $f$", or ``$f$ of $a$ equals $b$". If the function is clear from the context, we may write $a\mapsto b$. We may also refer to $a$ as an \textbf{input} of $f$ and $f(a)$ (or $b$) as an \textbf{output} of $f$. In this case, the domain of a function is the set of inputs while the range is the set of outputs for the function.
 
Notice that we can interpret our definition of function in terms of existence and uniqueness.  That is, $f:X\to Y$ is a function provided:
\begin{enumerate}
\item (Existence) For each $x\in X$, there exists $y\in Y$ such that $y=f(x)$, and
\item (Uniqueness) If $f(x)=y_1$ and $f(x)=y_2$, then $y_{1}=y_{2}$.
\end{enumerate}
In other words, every element of the domain is utilized and is utilized exactly once. However, there are no restrictions on whether an element of the codomain ever appears in the second coordinate of an ordered pair in the relation.  Yet if an element of $Y$ is in the range of $f$, it may appear in more than one ordered pair in the relation.

It follows immediately from the definition of function that two functions are equal if and only if they have the same domain, same codomain, and the same set of ordered pairs in the relation. That is, functions $f$ and $g$ are equal if and only if $\dom(f)=\dom(g)$, $\codom(f)=\codom(g)$, and $f(x)=g(x)$ for all $x\in X$.

Since functions are special types of relations, we can represent them using digraphs and graphs when practical. Digraphs for functions are often called \textbf{function} (or \textbf{mapping}) \textbf{diagrams}. When drawing function diagrams, it is standard practice to put the vertices for the domain on the left and the vertices for the codomain on the right, so that all directed edges point from left to right. We may also draw an additional arrow labeled by the name of the function from the domain to the codomain.

\begin{example}\label{ex:function}
Let $X=\{a,b,c,d\}$ to $Y=\{1,2,3,4\}$ and define the relation $f$ from $X$ to $Y$ via
\[
f=\{(a,2),(b,4),(c,4),(d,1)\}.
\]
Since each element $X$ appears exactly once as a first coordinate, $f$ is a function with domain $X$ and codomain $Y$ (i.e., $f:X\to Y$). In this case, we see that $\range(f)=\{1,2,4\}$.  Moreover, we can write things like $f(a)=2$ and $c\mapsto 4$, and say things like ``$f$ maps $b$ to 4" and ``the image of $d$ is 1."  The function diagram for $f$ is depicted in Figure~\ref{fig:function diagram}.
\end{example}

\tikzset{every loop/.style={min distance=8mm}}
\tikzstyle{vert} = [circle, draw, fill=grey,inner sep=0pt, minimum size=5mm]
\tikzstyle{b} = [draw,-stealth]

\begin{figure}[h!]
\begin{center}
\begin{tikzpicture}[->,>=stealth,shorten >=1pt,auto,node distance=2cm,semithick,scale=1.2]

\node[vert] (A) at (0,4) {\scriptsize $a$};
\node[vert] (B) at (0,3) {\scriptsize $b$};
\node[vert] (C) at (0,2) {\scriptsize $c$};
\node[vert] (D) at (0,1) {\scriptsize $d$};

\node[vert] (1) at (2.5,4) {\scriptsize $1$};
\node[vert] (2) at (2.5,3) {\scriptsize $2$};
\node[vert] (3) at (2.5,2) {\scriptsize $3$};
\node[vert] (4) at (2.5,1) {\scriptsize $4$};

\path (A) edge node {} (2)
(B) edge node {} (4)
(C) edge node {} (4)
(D) edge node {} (1);

\node[draw,fit=(A) (B) (C) (D),minimum width=1.5cm,gray] {};
\node[draw,fit=(1) (2) (3) (4),minimum width=1.5cm,gray] {};
\node (X) at (0,4.7) {$X$};
\node (Y) at (2.5,4.7) {$Y$};
\path (X) edge [bend left] node {$f$} (Y);
\end{tikzpicture}
\caption{Function diagram for a function from $X=\{a,b,c,d,\}$ to $Y=\{1,2,3,4\}$.}\label{fig:function diagram}
\end{center}
\end{figure}

\begin{problem}
Determine whether each of the relations defined in the following examples and problems is a function.
\begin{enumerate}[label=\textrm{(\alph*)}]
\item Example~\ref{ex:relation finite to finite} (see Figure~\ref{fig:digraph finite to finite})
\item Example~\ref{ex:circle} (see Figure~\ref{fig:unit circle})
\item Problem~\ref{prob:parabola}
\item Problem~\ref{prob:facebook}
\end{enumerate}
\end{problem}

\begin{problem}\label{prob:lots of potential functions}
Let $X=\{\circ, \square,\triangle,\smiley\}$ and $Y=\{a,b,c,d,e\}$. For each of the following relations, draw the corresponding digraph and determine whether the relation represents a function from $X$ to $Y$, $Y$ to $X$, $X$ to $X$, or does not represent a function. If the relation is a function, determine the domain, codomain, and range.
\begin{enumerate}[label=\textrm{(\alph*)}]
\item $f=\{(\circ, a),(\square,b),(\triangle,c),(\smiley,d)\}$
\item $g=\{(\circ, a),(\square,b),(\triangle,c),(\smiley,c)\}$
\item $h=\{(\circ, a),(\square,b),(\triangle,c),(\circ,d)\}$
\item $k=\{(\circ, a),(\square,b),(\triangle,c),(\smiley,d),(\square,e)\}$
\item $l=\{(\circ, e),(\square,e),(\triangle,e),(\smiley,e)\}$
\item $m=\{(\circ, a),(\triangle,b),(\smiley,c)\}$
\item \label{prob:identity} $i=\{(\circ,\circ),(\square, \square),(\triangle,\triangle),(\smiley,\smiley)\}$
\item\label{prob:happy} Define the relation $\operatorname{happy}$ from $Y$ to $X$ via $(y,\smiley)\in\operatorname{happy}$ for all $y\in Y$.
\item\label{prob:nugget} $\operatorname{nugget}=\{(\circ,\circ),(\square, \square),(\triangle,\triangle),(\smiley,\square)\}$
\end{enumerate}
\end{problem}

The last two parts of the previous problem make it clear that functions may have names consisting of more than one letter.  The function names $\sin$, $\cos$, $\log$, and $\ln$ are instances of this that you have likely encountered in your previous experience in mathematics. One thing that you may have never noticed is the type of font that we use for function names. It is common to italicize generic function names like $f$ but not common function names like $\sin$.  However, we always italicize the variables used to represent the input and output for a function. For example, consider the font types used in the expressions $\sin(x)$ and $\ln(a)$.

\begin{problem}
What properties does the digraph for a relation from $X$ to $Y$ need to have in order for it to represent a function?
\end{problem}

\begin{problem}
In high school I am sure that you were told that a graph represents a function if it passes the \textbf{vertical line test}.  Carefully state what the vertical line test says and then explain why it works.
\end{problem}

Sometimes we can define a function using a formula. For example, we can write $f(x)=x^2-1$ to mean that each $x$ in the domain of $f$ maps to $x^2-1$ in the codomain. However, notice that providing only a formula is ambiguous!  A function is determined by its domain, codomain, and the correspondence between these two sets. If we only provide a description for the correspondence, it is not clear what the domain and codomain are.  Two functions that are defined by the same formula, but have different domains or codomains are \emph{not} equal.  

\begin{example}
The function $f:\mathbb{R}\to \mathbb{R}$ defined via $f(x)=x^{2}-1$ is not equal to the function $g:\mathbb{N}\to\mathbb{R}$ defined by $g(x)=x^{2}-1$ since the two functions do not have the same domain.
\end{example}

Sometimes we rely on context to interpret the domain and codomain.  For example, in a calculus class, when we describe a function in terms of a formula, we are implicitly assuming that the domain is the largest allowable subset of $\mathbb{R}$---sometimes called the \textbf{default domain}---that makes sense for the given formula while the codomain is $\mathbb{R}$. 

\begin{example}
If we write $f(x)=x^2-1$, $g(x)=\sqrt{x}$, and $h(x)=\frac{1}{x}$ without mentioning the domains, we would typically interpret these as the functions $f:\mathbb{R}\to \mathbb{R}$, $g:[0,\infty)\to \mathbb{R}$, and $h:\mathbb{R}\setminus \{0\}\to \mathbb{R}$ that are determined by their respective formulas.
\end{example}

\begin{problem}
Provide an example of each of the following.  You may draw a function diagram, write down a list of ordered pairs, or a write a formula as long as the domain and codomain are clear.
\begin{enumerate}[label=\textrm{(\alph*)}]
\item A function $f$ from a set with 4 elements to a set with 3 elements such that $\range(f)=\codom(f)$.
\item A function $g$ from a set with 4 elements to a set with 3 elements such that $\range(g)$ is strictly smaller than $\codom(g)$.
\end{enumerate}
\end{problem}

\begin{problem}
Let $f:X\to Y$ be a function and suppose that $X$ and $Y$ are finite sets with $n$ and $m$ elements, respectively, such that $n<m$.  Is it possible for $\range(f)=\codom(f)$?  If so, provide an example.  If this is not possible, explain why.
\end{problem}

There are a few special functions that we should know the names of.

\begin{definition}
If $X$ is a nonempty set, then the function $i_X:X\to X$ defined via $i_X(x)=x$ is called the \textbf{identity function on $X$}.
\end{definition}

\begin{example}
The relation defined in Problem~\ref{prob:lots of potential functions}\ref{prob:identity} is the identity function on $X=\{\circ, \square,\triangle,\smiley\}$.
\end{example}

\begin{problem}
Draw a portion of the graph of the identity function on $\mathbb{R}$ as a subset of $\mathbb{R}^2$.
\end{problem}

\begin{definition}
Any function $f:X\to Y$ defined via $f(x)=c$ for a fixed $c\in Y$ is called a \textbf{constant function}.
\end{definition}

\begin{example}
The function defined in Problem~\ref{prob:lots of potential functions}\ref{prob:happy} is an example of a constant function.  Notice that we can succinctly describe this function using the formula $\operatorname{happy}(y)=\smiley$.
\end{example}

\begin{definition}
A \textbf{piecewise-defined function} (or \textbf{piecewise function}) is a function defined by specifying its output on a partition of the domain.
\end{definition}

Note that ``piecewise" is a way of expressing the function, rather than a property of the function itself.

\begin{example}
We can express the function in Problem~\ref{prob:lots of potential functions}\ref{prob:nugget} as a piecewise function using the formula
\[
\operatorname{nugget}(x)=\begin{cases}
x, & \mbox{if } x\mbox{ is a geometric shape},\\
\square, & \mbox{otherwise}.
\end{cases}
\]
\end{example}

\begin{example}
The function $f:\mathbb{R}\to \mathbb{R}$ defined via
\[
f(x)=\begin{cases}
x^2-1, & \mbox{if } x\geq 0,\\
17, & \mbox{if } -2\leq x< 0,\\
-x, & \mbox{if } x<-2
\end{cases}
\]
is an example of a piecewise-defined function.
\end{example}

\begin{problem}
Define $f:\mathbb{R}\setminus\{0\}\to \mathbb{R}$ via $f(x)=\frac{|x|}{x}$. Express $f$ as a piecewise function.
\end{problem}

It is important to point out that not every function can be described using a formula! Despite your prior experience, functions that can be represented succinctly using a formula are rare.

We now turn our attention to a couple of properties that a function may or may not possess.

\begin{definition}
Let $f:X\to Y$ be a function.
\begin{enumerate}[label=\textrm{(\alph*)}]
\item The function $f$ is said to be \textbf{injective} (or \textbf{one-to-one}) if for all $y\in \range(f)$, there is a unique $x\in X$ such that $y=f(x)$.
\item The function $f$ is said to be \textbf{surjective} (or \textbf{onto}) if for all $y\in Y$, there exists $x\in X$ such that $y=f(x)$.
\item If $f$ is both injective and surjective, we say that $f$ is \textbf{bijective}.
\end{enumerate}
\end{definition}

An injective function is also be called an \textbf{injection}, a surjective function is called a \textbf{surjection}, and a bijective function is called a \textbf{bijection} (or a \textbf{one-to-one correspondence}). To prove that a function $f:X\to Y$ is an injection, we must prove that if $f(x_{1})=f(x_{2})$, then $x_{1}=x_{2}$. To show that $f$ is surjective, you should start with an arbitrary $y\in Y$ and then work to show that there exists $x\in X$ such that $y=f(x)$.

%To do this, start by assuming that $f(x_{1})=f(x_{2})$ and then work to show that $x_{1}=x_{2}$.  That is, a function $f$ is one-to-one if and only if for all $x_{1}, x_{2}\in X$, if $f(x_{1})=f(x_{2})$, then $x_{1}=x_{2}$. To show that $f$ is onto, you should start with an arbitrary $y\in Y$ and then work to show that there exists $x\in X$ such that $y=f(x)$.

The next two skeleton proofs are special cases of Skeleton Proof~\ref{skeleton:uniqueness} and Skeleton Proof~\ref{skeleton:existence}, respectively.

\begin{skeleton}[Proof that a function is injective]
Here is the general structure for proving that a function is injective.
\begin{center}
\framebox{
\begin{minipage}{6in}
\vspace{.1in}
\begin{proof}
Assume $f:X\to Y$ is a function defined by (or satisfying)\ldots \emph{[Use the given definition (or describe the given property) of $f$]}.  Let $x_1,x_2\in X$ and suppose $f(x_1)=f(x_2)$.
\begin{center}
$\ldots$ \ \emph{[Use the definition (or property) of $f$ to verify that $x_1=x_2$]} \ $\ldots$
\end{center}
\noindent Therefore, the function $f$ is injective.
\end{proof}
\end{minipage}
}
\end{center}
\end{skeleton}

\begin{skeleton}[Proof that a function is surjective]
Here is the general structure for proving that a function is surjective.
\begin{center}
\framebox{
\begin{minipage}{6in}
\vspace{.1in}
\begin{proof}
Assume $f:X\to Y$ is a function defined by (or satisfying)\ldots \emph{[Use the given definition (or describe the given property) of $f$]}.  Let $y\in Y$.
\begin{center}
$\ldots$ \ \emph{[Use the definition (or property) of $f$ to find some $x\in X$ such that $f(x)=y$]} \ $\ldots$
\end{center}
\noindent Therefore, the function $f$ is surjective.
\end{proof}
\end{minipage}
}
\end{center}
\end{skeleton}

\begin{problem}
Assume that $X$ and $Y$ are finite sets. Provide an example of each of the following.  You may draw a function diagram, write down a list of ordered pairs, or a write a formula as long as the domain and codomain are clear. 
\begin{enumerate}[label=\textrm{(\alph*)}]
\item A function $f:X\to Y$ that is injective but not surjective.
\item A function $f:X\to Y$ that is surjective but not injective.
\item A function $f:X\to Y$ that is a bijection.
\item A function $f:X\to Y$ that is neither injective nor surjective.
\end{enumerate}
\end{problem}

\begin{problem}
Provide an example of each of the following.  You may either draw a graph or write down a formula.  Make sure you have the correct domain.
\begin{enumerate}[label=\textrm{(\alph*)}]
\item A function $f:\mathbb{R}\to \mathbb{R}$ that is injective but not surjective.
\item A function $f:\mathbb{R}\to \mathbb{R}$ that is surjective but not injective.
\item A function $f:\mathbb{R}\to \mathbb{R}$ that is a bijection.
\item A function $f:\mathbb{R}\to \mathbb{R}$ that is neither injective nor surjective.
\end{enumerate}
\end{problem}

\begin{problem}
Fill in the blank with the appropriate word.
\begin{quote}
A function $f:X\to Y$ is \underline{\ \ \ \ \ \ \ \  \ \ \ \ \ \ \ \ \ } if and only if every horizontal line hits the graph of $f$ at most once.
\end{quote}
This statement is often called the \textbf{horizontal line test}.  Explain why the horizontal line test is true.
\end{problem}

\begin{problem}\label{prob:injective surjective functions}
Determine whether each of the following functions is injective, surjective, both, or neither.  In each case, you should provide a proof or a counterexamples as appropriate.
\begin{enumerate}[label=\textrm{(\alph*)}]
\item $f:\mathbb{R}\to \mathbb{R}$ defined via $f(x)=x^{2}$
\item $g:\mathbb{R}\to [0,\infty)$ defined via $g(x)=x^{2}$
\item $h:\mathbb{R}\to \mathbb{R}$ defined via $h(x)=x^{3}$
\item $k:\mathbb{R}\to \mathbb{R}$ defined via $k(x)=x^{3}-x$
\item\label{circles} $c: \mathbb{R}\times \mathbb{R}\to \mathbb{R}$ defined via $l(x,y)=x^{2}+y^{2}$
\item $N:\mathbb{N}\to \mathbb{N}\times \mathbb{N}$ defined via $N(n)=(n,n)$
\end{enumerate}
\end{problem}

\begin{problem}
Compare and contrast the definition of ``function" with the definition of ``injective function". Consider the vertical line test and horizontal line test in your discussion.  Moreover, attempt to capture what it means for a relation to not be a function and for a function to not be an injection by drawing portions of a digraph.
\end{problem}

The next theorem should come as no surprise.

\begin{theorem}
The identity function $i_X$ on a nonempty set $X$ is a bijection.
\end{theorem}

\begin{problem}
Let $A$ and $B$ be nonempty sets and let $S$ be a nonempty subset of $A\times B$.  Define $\pi_{1}:S\to A$ and $\pi_{2}:S\to B$ via $\pi_{1}(a,b)=a$ and $\pi_{2}(a,b)=b$.  We call $\pi_{1}$ and $\pi_{2}$ the \textbf{projections} of $S$ onto $A$ and $B$, respectively.
\begin{enumerate}[label=\textrm{(\alph*)}]
\item Provide examples to show that $\pi_{1}$ does not need to be injective nor surjective.
\item Suppose that $S$ is also a function. Is $\pi_{1}$ injective? Is $\pi_{1}$ surjective?  How about $\pi_{2}$?
\end{enumerate}
\end{problem}

The next theorem says that if we have an equivalence relation on a nonempty set, the mapping that assigns each element to its respective equivalence class is a surjective function.  

\begin{theorem}\label{thm:canonical projection}
If $\sim$ is an equivalence relation on a nonempty $A$, then the function $\phi:A\to A/\mathord\sim$ defined via $\phi(x)=[x]$ is a surjection.
\end{theorem}

The function from the previous theorem is sometimes called the \textbf{canonical projection} induced by $\sim$.

\begin{problem}
Under what circumstances would the function from the previous theorem also be injective?
\end{problem}

Let's explore whether we can weaken the hypotheses of Theorem~\ref{thm:canonical projection}. 

\begin{problem}
Let $R$ be a relation on a nonempty set $A$.
\begin{enumerate}[label=\textrm{(\alph*)}]
\item What conditions on $R$ must hold in order for $f:A\to \Rel(R)$ defined via $f(a)=\rel(a)$ to be a function?
\item What additional conditions, if any, must hold on $R$ in order for $f$ to be a surjective function?
\end{enumerate}
\end{problem}

\begin{problem}
Let $A$ be a nonempty set.
\begin{enumerate}[label=\textrm{(\alph*)}]
\item Suppose $R$ is an equivalence relation on $A$. Under what conditions is $R$ a function from $A$ to $A$?
\item Suppose $f:A\to A$ is a function. Under what condition is $f$ an equivalence relation on $A$?
\end{enumerate}
\end{problem}

The next problem illustrates that some care must be taken when attempting to define a function.

\begin{problem}\label{prob:not well defined}
For each of the following, explain why the given description does not define a function.
\begin{enumerate}[label=\textrm{(\alph*)}]
\item Define $f:\{1,2,3\}\to \{1,2,3\}$ via $f(a)=a-1$.
\item Define $g:\mathbb{N}\to \mathbb{Q}$ via $g(n)=\frac{n}{n-1}$.
\item Let $A_1=\{1,2,3\}$ and $A_2=\{3,4,5\}$. Define $h:A\cup B\to \{1,2\}$ via
\[
h(x)=\begin{cases}
1, & \mbox{if }x\in A_1\\
2, & \mbox{if }x\in A_2.
\end{cases}
\]
\item Define $s:\mathbb{Q}\to \mathbb{Z}$ via $s(a/b)=a+b$.
\end{enumerate}
\end{problem}

In mathematics, we say that an expression is \textbf{well defined} (or \textbf{unambiguous}) if its definition yields a unique interpretation.  Otherwise, we say that the expression is not well defined (or is \textbf{ambiguous}).  For example, if $a,b,c\in\mathbb{R}$, then the expression $abc$ is well defined since it does not matter if we interpret this as $(ab)c$ or $a(bc)$ (i.e., the real numbers are associative under multiplication).  This issue was lurking behind the scenes in the statement of Theorem~\ref{thm:modular sums products}.  In particular, the expressions
\[
[a_1]_n+[a_2]_n+\cdots+ [a_k]_n
\]
and
\[
[a_1]_n [a_2]_n \cdots  [a_k]_n
\]
are well defined in $\mathbb{Z}/n\mathbb{Z}$ in light of Theorems~\ref{thm:modular add comm assoc}\ref{modular add assoc} and \ref{thm:modular mult comm assoc}\ref{modular mult assoc}.

When we attempt to define a function, it may not be clear without doing some work that our definition really does yield a function. If there is some potential ambiguity in the definition of a function that ends up not causing any issues, we say that the function is well defined. However, this phrase is a bit of misnomer since all functions are well defined. The issue of whether a description for a proposed function is well defined often arises when defining things in terms of representatives of equivalence classes, or more generally in terms of how an element of the domain is written.  For example, the descriptions given in parts (c) and (d) of Problem~\ref{prob:not well defined} are not well defined.  To show that a potentially ambiguous description for a function $f:X\to Y$ is well defined prove that if $a$ and $b$ are two representations for the same element in $X$, then $f(a)=f(b)$.

\begin{problem}
For each of the following, determine whether the description determines a well-defined function.
\begin{enumerate}[label=\textrm{(\alph*)}]
\item Define $f:\mathbb{Z}/5\mathbb{Z}\to \mathbb{N}$ via
\[
f([a]_5)=\begin{cases}
0, & \mbox{if } a\mbox{ is even}\\
1, & \mbox{if } a\mbox{ is odd}.
\end{cases}
\]
\item Define $g:\mathbb{Z}/6\mathbb{Z}\to \mathbb{N}$ via
\[
g([a]_6)=\begin{cases}
0, & \mbox{if } a\mbox{ is even}\\
1, & \mbox{if } a\mbox{ is odd}.
\end{cases}
\]
\end{enumerate}
\end{problem}

Given any function, we can define an equivalence relation on its domain, where the equivalence classes correspond to the elements that map to the same element of the range.

\begin{theorem}\label{thm:equiv relation from inverse images}
Let $f:X\to Y$ be a function and define $\sim$ on $X$ via $a\sim b$ if and only if $f(a) = f(b)$. Then $\sim$ is an equivalence relation on $X$.
\end{theorem}

It follows immediately from Theorem~\ref{thm:equiv yields partition2} that the equivalence classes induced by the equivalence relation in Theorem~\ref{thm:equiv relation from inverse images} partition the domain of a function.

\begin{problem}\label{prob:equiv classes from inverse images}
For each of the functions defined in the following example and problem, identify the equivalence classes induced by the equivalence relation from Theorem~\ref{thm:equiv relation from inverse images} for the given function.
\begin{enumerate}[label=\textrm{(\alph*)}]
\item Example~\ref{ex:function}
\item Problem~\ref{prob:injective surjective functions}\ref{circles}
\end{enumerate}
\end{problem}

If $f$ is a function, the equivalence relation in Theorem~\ref{thm:equiv relation from inverse images} allows us to construct a bijective function whose domain is the set of equivalence classes and whose range coincides with the range of $f$.  This is an important idea that manifests itself in many areas of mathematics.  One such instance is the First Isomorphism Theorem for Groups, which is a fundamental theorem in a branch of mathematics called group theory.  When proving the following theorem, the first thing you should do is verify that the description for $g$ is well defined.

\begin{theorem}\label{thm:iso theorem for sets}
Let $f:X\to Y$ be a function and define $\sim$ on $X$ via $a\sim b$ if and only if $f(a) = f(b)$. Then the function $\overline{f}:X/\mathord\sim\to \range(f)$ defined via $\overline{f}([a]) = f(a)$ is a bijection.
\end{theorem}

\begin{problem}
Consider the equivalence classes you identified in parts (a) and (b) of Problem~\ref{prob:equiv classes from inverse images}.
\begin{enumerate}[label=\textrm{(\alph*)}]
\item Draw the function diagram for the function $\overline{f}$ as defined in Theorem~\ref{thm:iso theorem for sets}, where $f$ is the function defined in Example~\ref{ex:function}.
\item Describe the function $\overline{c}$ as defined in Theorem~\ref{thm:iso theorem for sets}, where $c$ is the function defined in Problem~\ref{prob:injective surjective functions}\ref{circles}.
\end{enumerate}
\end{problem}

\end{section}