\documentclass[12pt,oneside]{book}

\usepackage[scale=2]{ccicons}
\usepackage{caption}
\usepackage[labelfont={}]{subcaption}
\captionsetup{subrefformat=parens}
\usepackage{booktabs}
\usepackage{epigraph}
\usepackage{enumitem}
\usepackage{newclude}
\usepackage{multicol}
\usepackage{tabu}
\usepackage[table]{xcolor}
\usepackage{tikz}
\usetikzlibrary{arrows,automata,positioning,fit,shapes}
\usepackage{rotating}
\usepackage[notextcomp]{kpfonts} 
\usepackage{graphicx}
\usepackage{eurosym}
\usepackage{amsfonts}
\usepackage{mathtools}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{wasysym}
\usepackage{amsthm}
\usepackage[margin=1in]{geometry}
\usepackage[hang,flushmargin]{footmisc}
\usepackage{color}
\usepackage[breaklinks]{hyperref}
\hypersetup{
	colorlinks=true,
	linkcolor=darkblue,
	anchorcolor=darkblue,
	citecolor=darkblue,
	pagecolor=darkblue,
	urlcolor=darkblue,
	pdftitle={},
	pdfauthor={},
    bookmarksnumbered
}

%%% Colors %%%

\definecolor{darkblue}{rgb}{0, 0, .6}
\definecolor{grey}{rgb}{.7, .7, .7}
\definecolor{naugreen}{cmyk}{.43,0,.34,.38}
\definecolor{naublue}{cmyk}{1,.72,0,.32}
\definecolor{mediterranean}{cmyk}{.67,0,.08,.3}
\definecolor{rose}{cmyk}{0,1.00,.20,0}
\definecolor{darkorchid}{cmyk}{.6,.9,0,.05}
\definecolor{butterfly}{cmyk}{.95,.59,0,.10}
\definecolor{springgreen}{cmyk}{1.00,0,.70,.02}
\definecolor{darkred}{cmyk}{0,1,1,.5}
\definecolor{nectarine}{cmyk}{0,0.70,1.00,0}
\definecolor{icyblue}{cmyk}{.84,.25,0,.06}
\definecolor{orange}{RGB}{255,102,0}
\definecolor{ggreen}{RGB}{0,153,0}
\definecolor{darkblue}{RGB}{0,0,255}
\definecolor{purple}{RGB}{153,51,255}
\definecolor{turq}{RGB}{72,209,204}
\definecolor{lightskyblue}{cmyk}{.4,.11,0,.2}
\definecolor{softplum}{cmyk}{.02,.45,0,.68}%.37
\definecolor{lightorange}{cmyk}{0,.45,.59,.02}
\definecolor{brickorange}{cmyk}{0,.59,.59,.20}
\definecolor{deeppink}{cmyk}{0,.6,.3,.3}
\definecolor{orchid}{cmyk}{0,.49,.02,.05}

\usepackage{fancyhdr}
\pagestyle{fancy}
\lhead{\leftmark}
\chead{}
\rhead{}
\lfoot{}
\cfoot{\thepage}
\rfoot{}

\theoremstyle{definition}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{acknowledgement}[theorem]{Acknowledgement}
\newtheorem{algorithm}[theorem]{Algorithm}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{case}[theorem]{Case}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conclusion}[theorem]{Conclusion}
\newtheorem{condition}[theorem]{Condition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{journal}[theorem]{Journal}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{solution}[theorem]{Solution}
\newtheorem{summary}[theorem]{Summary}
\newtheorem{skeleton}[theorem]{Skeleton Proof}
\newtheorem{activity}[theorem]{Activity}
\newtheorem{intuitivedef}[theorem]{Intuitive Definition}
\newtheorem{question}[theorem]{Question}
\newtheorem{fieldaxioms}[theorem]{Field Axioms}
\newtheorem{orderaxioms}[theorem]{Order Axioms}
\newtheorem{completenessaxiom}[theorem]{Completeness Axiom}
\newtheorem{additionalaxioms}[theorem]{Additional Order Axioms}

\newsavebox{\savepar}
\newenvironment{textbox}{\noindent\begin{lrbox}{\savepar}\begin{minipage}[c]{.98\textwidth}}{\end{minipage}\end{lrbox}\fcolorbox{black}{white}{\usebox{\savepar}}}

\newcommand{\dom}{\operatorname{Dom}}
\newcommand{\codom}{\operatorname{Codom}}
\newcommand{\range}{\operatorname{Rng}}
\newcommand{\lcm}{\operatorname{lcm}}
\newcommand{\card}{\operatorname{card}}
\newcommand{\rel}{\operatorname{rel}}
\newcommand{\Rel}{\operatorname{Rel}}

\begin{document}

\title{An Introduction to Proof via \\Inquiry-Based Learning}
\author{Dana C.~Ernst, PhD\\
Northern Arizona University}
\date{Version Fall 2021}

\maketitle
\thispagestyle{empty}

\noindent\copyright{ \the\year\ Dana C.~Ernst.  Some Rights Reserved.\\

\bigskip

\noindent This book is intended to be a task sequence for an introduction to proof course that utilizes an inquiry-based learning (IBL) approach.  You can find the most up-to-date version of this textbook on GitHub:
\begin{center}
\url{http://dcernst.github.io/IBL-IntroToProof/}
\end{center}
I would be thrilled if you used this textbook and improved it. If you make any modifications, you can either make a pull request on GitHub or submit the improvements via email.  You are also welcome to fork the source and modify the text for your purposes as long as you maintain the license below.

\bigskip

\noindent This work is licensed under the Creative Commons Attribution-Share Alike 4.0 United States License.  You may copy, distribute, display, and perform this copyrighted work, but only if you give credit to Dana C.~Ernst, and all derivative works based upon it must be published under the Creative Commons Attribution-Share Alike 4.0 International License. Please attribute this work to Dana C.~Ernst, Mathematics Faculty at Northern Arizona University, \url{dana.ernst@nau.edu}, as well as the individuals listed below. To view a copy of this license, visit
\begin{center}
\url{https://creativecommons.org/licenses/by-sa/4.0/}
\end{center}
or send a letter to Creative Commons, 171 Second Street, Suite 300, San Francisco, California, 94105, USA.}

\medskip

\begin{center}
\ccbysa
\end{center}

\bigskip

\noindent Below is a partial list of people (alphabetical by last name) that I need to thank for supplying content, advice, and feedback.
\begin{itemize}
\item \href{https://thalestriangles.blogspot.com}{Joshua Bowman} (Pepperdine University). The Preface is a modified version of Joshua's \emph{An IBL preface}.
\item \href{https://math.depaul.edu/cdrupies/}{Chris Drupieski} (DePaul University), \href{http://math.depaul.edu/tpeter21/}{T.~Kyle Petersen} (DePaul University), and \href{http://math.depaul.edu/bridget/}{Bridget Tenner} (DePaul University). Modifications that these three made to the book inspired me to streamline some of the exposition, especially in the early chapters.
\item \href{http://www.paulellis.org}{Paul Ellis} (Manhattanville College). Paul has provided lots of useful feedback and several suggestions for improvements. Paul suggested problems for Chapter~\ref{chap:Induction} and provided an initial draft of Section~\ref{sec:Images and Inverse Images}: Images and Inverse Images.
\item \href{http://jasongrout.org}{Jason Grout} (Bloomberg, L.P.).  I am extremely grateful to Jason for feedback on early versions of this manuscript, as well as for helping me with a variety of technical aspects of writing an open-source textbook.
\item \href{http://home.snc.edu/andershendrickson/}{Anders Hendrickson} (St.~Norbert College). Anders is the original author of the content in Appendix~\ref{appendix:elements_of_style}: Elements of Style for Proofs. The current version in Appendix~\ref{appendix:elements_of_style} is a result of modifications made by myself with some suggestions from David Richeson.
\item \href{http://www.hsc.edu/rebecca-jayne}{Rebecca Jayne} (Hampden--Sydney College). The current version of Section~\ref{sec:CompleteInduction}: Complete Induction is a derivative of content contributed by Rebecca.
\item \href{http://www4.csudh.edu/library/info/civic-directory/f-j/matthew-g-jones}{Matthew Jones} (CSU Dominguez Hills) and \href{http://www.stanyoshinobu.com}{Stan Yoshinobu} (University of Toronto). A few of the sections were originally adaptations of notes written by Matt and Stan. Early versions of this textbook relied heavily on their work. Moreover, Matt and Stan were two of the key players that contributed to shaping my approach to teaching.
\item \href{https://embracinglifewithmajorrevisions.org/2017/07/12/rights-of-the-learner-an-introduction/}{Crystal Kalinec-Craig} (University of Texas at San Antonio). Section~\ref{sec:Rights of the Learner}: Rights of the Learner is an adaptation of a similar list written by Crystal.
\item \href{http://users.dickinson.edu/~richesod/}{David Richeson} (Dickinson College). David is responsible for much of the content in Appendix~\ref{appendix:fancy_math_terms}: Fancy Mathematical Terms and Appendix~\ref{appendix:definitions}: Definitions in Mathematics. In addition, the current version of Chapter~\ref{chap:three famous theorems}: Three Famous Theorems is heavily based on content contributed by David.
\item \href{http://www2.kenyon.edu/Depts/Math/schumacherc/public_html/}{Carol Schumacher} (Kenyon College). When I was transitioning to an IBL approach to teaching, Carol was one of my mentors and played a significant role in my development as a teacher.  Moreover, this work is undoubtably influenced my Carol's excellent book \emph{Chapter Zero: Fundamental Notions of Advanced Mathematics}, which I used when teaching my very first IBL course.
\item \href{http://webpages.csus.edu/wiscons/}{Josh Wiscons} (CSU Sacramento). The current version of Section~\ref{sec:ModularArithmetic}: Modular Arithmetic is a derivative of content contributed by Josh.
\end{itemize}

\tableofcontents
%\thispagestyle{empty}

\include{Preface}
\include{Introduction}
\include*{IntroToMath}
\include*{TasteNumberTheory}
\include*{IntroToLogic}
\include*{NegatingAndContradiction}
\include*{IntroQuantification}
\include*{MoreQuantification}
\include*{SetTheory}
\include*{Sets}
\include*{RussellsParadox}
\include*{PowerSets}
\include*{IndexingSets}
\include*{CartesianProducts}
\include*{RealNumbers}
\include*{AxiomsRealNumbers}
\include*{Topology}
\include*{Induction}
\include*{IntroInduction}
\include*{MoreInduction}
\include*{CompleteInduction}
\include*{ThreeFamousTheorems}
\include*{FundamentalTheoremArithmetic}
\include*{IrrationalityRoot2}
\include*{InfinitudeOfPrimes}
\include*{RelationsPartitions}
\include*{Relations}
\include*{EquivalenceRelations}
\include*{Partitions}
\include*{ModularArithmetic}
%\include*{Posets}
\include*{Functions}
\include*{IntroFunctions}
\include*{InjectiveSurjectiveFunctions} 
\include*{CompositionsInverses}
\include*{ImagesInverseImages}
\include*{ContinuousFunctions}
\include*{Cardinality}
\include*{IntroCardinality}
\include*{FiniteSets}
\include*{InfiniteSets}
\include*{CountableSets}
\include*{UncountableSets}
%Appendices
\include{ElementsOfStyle}
\include{FancyMathematicalTerms}
\include{Paradoxes}
\include{Definitions}

\end{document}
