\chapter{Introduction}\label{chap:intro}

\epigraphhead[70pt]{
\epigraph{The mathematician does not study pure mathematics because it is useful; he studies it because he delights in it, and he delights in it because it is beautiful.}{Henri Poincar\'e, mathematician \& physicist}}

\begin{section}{What is This Book All About?}%new title?

This book is intended to be used for a one-semester/quarter introduction to proof course (sometimes referred to as a transition to proof course). The intended audience is mathematics majors and minors. The book is designed around an inquiry-based learning (IBL) approach to teaching and learning mathematics, but makes no assumptions about the specifics of how the instructor chooses to implement this approach.  Loosely speaking, IBL is a student-centered method of teaching that engages students in sense-making activities.  Students are given tasks requiring them to solve problems, conjecture, experiment, explore, create, and communicate.  Rather than showing facts or a clear, smooth path to a solution, the instructor guides and mentors students via well-crafted problems through an adventure in mathematical discovery.  According to \href{https://www.colorado.edu/eer/sites/default/files/attached-files/laursenrasmussencommentaryauthorversion0219.pdf}{Laursen and Rasmussen (2019)}, the Four Pillars of IBL are:
\begin{itemize}
\item Students engage deeply with coherent and meaningful mathematical tasks.
\item Students collaboratively process mathematical ideas.
\item Instructors inquire into student thinking.
\item Instructors foster equity in their design and facilitation choices.
\end{itemize}

The book includes more content than one can expect to cover in a single semester. This allows the instructor to pick and choose the sections that suit their needs and desires. Each chapter takes a focused approach to the included topics, but also includes many gentle exercises aimed at developing intuition.

Up to this point, it is likely that your experience of mathematics has been about using formulas and algorithms. You are used to being asked to do things like: ``solve for x", ``take the derivative of this function", ``integrate this function", etc. Accomplishing tasks like these usually amounts to mimicking examples that you have seen in class or in your textbook. This is only one part of mathematics. Mathematicians do much more than just use formulas.  Mathematicians experiment, make conjectures, write definitions, and prove theorems.  While engaging with the material contained in this book, we will learn about doing all of these things. Ultimately, the focus of this book is on producing and discovering mathematics. Engaging in this endeavor may be a shock t to your system.

Just like learning to play an instrument or sport, you will have to learn new skills and ideas. Along this journey, you should expect a cycle of victory and defeat, experiencing a full range of emotions.  Sometimes you will feel exhilarated, other times you might be seemingly paralyzed by extreme confusion. Learning a new skill requires dedication and patience during periods of frustration. As the author of this text, I am here to guide and challenge you, but I cannot do the learning for you, just as a music teacher cannot move your fingers and your heart for you.  Only you can do that.  

Do your best, be prepared to put in a lot of time, and do all the work. When you work hard and you come to understand, you feel good about yourself.  In the meantime, you have to believe that your work will pay off in intellectual development.




You will experience a full range of emotions, sometimes feeling  exhilarated, other times seemingly paralyzed by extreme frustration.  You should expect a cycle of victory and defeat, 

Figuring it out, victories, defeats, and all that is part of real life is what you can expect.  Most importantly it will be rewarding.  Learning mathematics requires dedication.  It will require that you be patient despite periods of confusion.  It will require that you persevere in order to understand.  

As the author of this text, I am here to guide you, but I cannot do the learning for you, just as a music teacher cannot move your fingers and your heart for you.  Only you can do that.  Do your best, be prepared to put in a lot of time, and do all the work. When you work hard and you come to understand, you feel good about yourself.  In the meantime, you have to believe that your work will pay off in intellectual development.

Your progress will be fueled by your ability to wrestle with mathematical ideas and to prove theorems.  As you work through the book, you will find that you have ideas for proofs, but you are unsure of them.  Do not be afraid to tinker and make mistakes.  You can always revisit your work as you become more proficient. Do not expect to do most things perfectly on your first attempt. 

This is a very exciting time in your mathematical career.  It is where you learn what mathematics is really about!

Ultimately, mathematicians are in the business of proving theorems. However,




Ultimately, mathematicians are in the business of proving theorems and this is exactly our endeavor. For the first time, you will be exposed to what doing mathematics is really all about. This will most likely be a shock to your system. 






The steps you take to solve problems like these are always justified by mathematical theorems.


, but rarely are you paying explicit attention to when you are actually using these facts. Furthermore, justifying (i.e., proving) the mathematical facts you use may have been omitted by the instructor. And, even if the instructor did prove a given theorem, you may not have taken the time or have been able to digest the content of the proof.

That is only one part of mathematics. Mathematicians do much more than just use formulas.  Mathematicians experiment, make conjectures, write definitions, and prove theorems.  While engaging with the material contained in this book, we will learn about doing all of these things. The focus of this book is on producing and discovering mathematics.  






This course will likely be different than any other math class that you have taken before for two main reasons. First, you are used to being asked to do things like: ``solve for x", ``take the derivative of this function", ``integrate this function", etc. Accomplishing tasks like these usually amounts to mimicking examples that you have seen in class or in your textbook. The steps you take to ``solve" problems like these are always justified by mathematical facts (theorems), but rarely are you paying explicit attention to when you are actually using these facts. Furthermore, justifying (i.e., proving) the mathematical facts you use may have been omitted by the instructor. And, even if the instructor did prove a given theorem, you may not have taken the time or have been able to digest the content of the proof.

Unlike previous courses, this course is all about ``proof". Mathematicians are in the business of proving theorems and this is exactly our endeavor. For the first time, you will be exposed to what doing mathematics is really all about. This will most likely be a shock to your system. 





Up to this point, it is likely that your experience of mathematics has been about using formulas and algorithms. That is only one part of mathematics. Mathematicians do much more than just use formulas.  Mathematicians experiment, make conjectures, write definitions, and prove theorems.  While engaging with the material contained in this book, we will learn about doing all of these things. The focus of this book is on producing and discovering mathematics.  

learning to \emph{do} mathematics. 
  
Just like learning to play an instrument or sport, you will have to learn new skills and ideas.  Sometimes you will feel good, sometimes frustrated.  You will probably go through a range of feelings from being exhilarated, to being stuck.  Figuring it out, victories, defeats, and all that is part of real life is what you can expect.  Most importantly it will be rewarding.  Learning mathematics requires dedication.  It will require that you be patient despite periods of confusion.  It will require that you persevere in order to understand.  As the author of this text, I am here to guide you, but I cannot do the learning for you, just as a music teacher cannot move your fingers and your heart for you.  Only you can do that.  Do your best, be prepared to put in a lot of time, and do all the work. When you work hard and you come to understand, you feel good about yourself.  In the meantime, you have to believe that your work will pay off in intellectual development.

Your progress will be fueled by your ability to wrestle with mathematical ideas and to prove theorems.  As you work through the book, you will find that you have ideas for proofs, but you are unsure of them.  Do not be afraid to tinker and make mistakes.  You can always revisit your work as you become more proficient. Do not expect to do most things perfectly on your first attempt. 

This is a very exciting time in your mathematical career.  It is where you learn what mathematics is really about!

\epigraph{Mathematics, rightly viewed, possesses not only truth, but supreme beauty---a beauty cold and austere, like that of sculpture, without appeal to any part of our weaker nature, without the gorgeous trappings of painting or music, yet sublimely pure, and capable of a stern perfection such as only the greatest art can show. The true spirit of delight, the exaltation, the sense of being more than Man, which is the touchstone of the highest excellence, is to be found in mathematics as surely as poetry.}{Bertrand Russell, philosopher \& mathematician}

\end{section}

\begin{section}{An Inquiry-Based Approach}


This course will likely be different than any other math class that you have taken before for two main reasons. First, you are used to being asked to do things like: ``solve for x", ``take the derivative of this function", ``integrate this function", etc. Accomplishing tasks like these usually amounts to mimicking examples that you have seen in class or in your textbook. The steps you take to ``solve" problems like these are always justified by mathematical facts (theorems), but rarely are you paying explicit attention to when you are actually using these facts. Furthermore, justifying (i.e., proving) the mathematical facts you use may have been omitted by the instructor. And, even if the instructor did prove a given theorem, you may not have taken the time or have been able to digest the content of the proof.

Unlike previous courses, this course is all about ``proof". Mathematicians are in the business of proving theorems and this is exactly our endeavor. For the first time, you will be exposed to what doing mathematics is really all about. This will most likely be a shock to your system. Considering the number of mathematics courses that you have taken before you arrived here, one would think that you have some idea what mathematics is all about. You must be prepared to modify your paradigm. The second reason why this course will be different for you is that the method by which the class will run and the expectations I have of you will be different. In a typical course, math or otherwise, you sit and listen to a lecture. (Hopefully) These lectures are polished and well-delivered. You may have often been lured into believing that the instructor has opened up your head and is pouring knowledge into it. I absolutely love lecturing and I do believe there is value in it, but I also believe that in reality most students do not learn by simply listening. You must be active in the learning you are doing. I am sure each of you have said to yourselves, ``Hmmm, I understood this concept when the professor was going over it, but now that I am alone, I am lost." In order to promote a more active participation in your learning, we will incorporate ideas from an educational philosophy called ...


This is not a typical textbook or one in which mimicking prefabricated examples will lead you to success. You will be expected to work actively to construct your own understanding of the topics at hand. Many of the concepts you learn and problems you work on will be new to you and ask you to stretch your thinking. You will experience \emph{frustration} and \emph{failure} before you experience \emph{understanding}. This is part of the normal learning process. If you are doing things well, you should be confused on a regular basis. The material is too rich for a human being to completely understand it immediately. Your viability as a professional in the modern workforce depends on your ability to embrace this learning process and make it work for you.

%In order to promote a more active participation in your learning, this book adheres to an educational philosophy called inquiry-based learning (IBL).  Loosely speaking, IBL is a student-centered method of teaching mathematics that engages students in sense-making activities.  Students are given tasks requiring them to solve problems, conjecture, experiment, explore, create, and communicate.  Rather than showing facts or a clear, smooth path to a solution, this book will guide and mentor you via well-crafted problems through an adventure in mathematical discovery. 

In many mathematics classrooms, ``doing mathematics" means following the rules dictated by the teacher, and ``knowing mathematics" means remembering and applying them. However, an inquiry-based-learning (IBL) approach challenges students to create or discover mathematics. IBL is a student-centered method of teaching mathematics that engages students in sense-making activities.  Students are given tasks requiring them to solve problems, conjecture, experiment, explore, create, and communicate.  Rather than showing facts or a clear, smooth path to a solution, this book will guide and mentor you via well-crafted problems through an adventure in mathematical discovery. Effective IBL courses come in a variety of forms, but they all possess a few essential ingredients. According to \href{https://www.colorado.edu/eer/sites/default/files/attached-files/laursenrasmussencommentaryauthorversion0219.pdf}{Laursen and Rasmussen (2019)}, the Four Pillars of IBL are:
\begin{itemize}
\item Students engage deeply with coherent and meaningful mathematical tasks.
\item Students collaboratively process mathematical ideas.
\item Instructors inquire into student thinking.
\item Instructors foster equity in their design and facilitation choices.
\end{itemize}

Moreover, students should---as much as possible---be responsible for guiding the acquisition of knowledge and validating the ideas presented. That is, students should not be looking to the instructor as the sole authority. In an IBL course, instructor and students have joint responsibility for the depth and progress of the course.

\textbf{Is this from Joshua Bowman???}
%%%
While progressing through this book, you will be asked to engage with the material and to produce mathematics. The best way to learn mathematics is by doing mathematics.  Someone cannot master a musical instrument or a martial art by simply watching, and in a similar fashion, you cannot master mathematics by simply watching; you must do mathematics!

In any act of creation, there must be room for experimentation, and thus allowance for mistakes, even failure. Mistakes are inevitable, and they should not be an obstacle to further progress. Productive struggle and mistakes provide opportunities for growth.  It is normal to struggle and be confused as you work through new material. Accepting that means you can keep working even while feeling stuck, until you overcome and reach even greater accomplishments.
%%%%

Furthermore, it is important to understand that solving genuine problems is difficult and takes time.  You should not expect to complete each problem in 10 minutes or less.  Sometimes you might have to stare at the problem for an hour before even understanding how to get started.

%from Kyle: I like to think of the difference between a lecture-driven classroom and an IBL classroom like the difference between taking a cable car to the summit of a mountain and climbing to the top of the same mountain. One could argue that both approaches can get you to the top of the mountain, but clearly the experiences are different. And when you come to a new mountain, for which there is no cable car, who is better prepared to reach the summit?

%also think about: Learning like a jungle tiger.

\epigraph{Don't fear failure.  Not failure, but low aim, is the crime. In great attempts it is glorious even to fail.}{Bruce Lee, martial artist \& actor}

\end{section}

\begin{section}{Rights of the Learner}\label{sec:Rights of the Learner}
As a reader of this textbook, you have the right to:
\begin{enumerate}
\item be confused,
\item make a mistake and to revise your thinking,
\item speak, listen, and be heard, and
\item enjoy doing mathematics.
\end{enumerate}

\epigraph{You may encounter many defeats, but you must not be defeated.}{Maya Angelou, poet \& activist}

\end{section}

\begin{section}{Structure of the Textbook}

As you read the textbook, you will be required to digest the material in a meaningful way.  It is your responsibility to read and understand new definitions and their related concepts.  In addition, you will be asked to complete problems aimed at solidifying your understanding of the material.  Most importantly, you will be asked to make conjectures, produce counterexamples, and prove theorems. All of these tasks will almost always be challenging.

The items labeled as \textbf{Definition} and \textbf{Example} are meant to be read and digested.  However, the items labeled as \textbf{Problem}, \textbf{Theorem}, \textbf{Corollary}, and \textbf{Lemma} require action on your part.  Items labeled as \textbf{Problem} are sort of a mixed bag. Some Problems are computational in nature and aimed at improving your understanding of a particular concept while others ask you to provide a counterexample for a statement if it is false or to provide a proof if the statement is true. Items with the \textbf{Theorem}, \textbf{Corollary}, and \textbf{Lemma} designation are mathematical facts and the intention is for you to produce a valid proof of the given statement.  A lemma is a minor result whose sole purpose is to help in proving a theorem. The main difference between a theorem and a corollary is that corollaries are typically statements that follow quickly from a previous theorem.  In general, you should expect corollaries to have very short proofs.  However, that does not mean that you cannot produce a more lengthy yet valid proof of a corollary.

Oftentimes, the problems and theorems are guiding you towards a substantial, more  general result. Other times, they are designed to get you to apply ideas in a new way. One thing to always keep in mind is that every task in this book can be done by you, the student. But it may not be on your first try, or even your second.

Discussion of new topics is typically kept at a minimum and there are very few examples in this book. This is intentional.  One of the objectives of the items labeled as \textbf{Problem} is for you to produce the examples needed to internalize unfamiliar concepts.  The ultimate goal of this book is to help you develop a deep and meaningful understanding of the processes of producing mathematics by putting you in direct contact with mathematical phenomena.

\epigraph{Don't just read it; fight it! Ask your own questions, look for your own examples, discover your own proofs. Is the hypothesis necessary? Is the converse true? What happens in the classical special case? What about the degenerate cases? Where does the proof use the hypothesis?}{Paul Halmos, mathematician}

\end{section}

\begin{section}{Some Minimal Guidance}\label{sec:minimal guidance}
Especially in the opening sections, it will not be clear what facts from your prior experience in mathematics we are ``allowed" to use.  Unfortunately, addressing this issue is difficult and is something we will sort out along the way.  However, in general, here are some minimal guidelines to keep in mind.

\begin{itemize}
\item The statement you are proving should be on the same page as the beginning of your proof.   
\item You should indicate where the proof begins by writing ``\emph{Proof.}" at the beginning.  
\item Make it clear to yourself and the reader what your assumptions are at the very beginning of your proof. Typically, these statements will start off ``Assume\ldots", ``Suppose\ldots", or ``Let\ldots".  Sometimes there will be some implicit assumptions that we can omit, but at least in the beginning, you should get in the habit of clearly stating your assumptions up front. 
\item Carefully consider the order in which you write your proof. Loosely speaking, each sentence should follow from an earlier sentence in your proof or possibly a result you have already proved.
\item Unlike the experience many of you had writing proofs in your high school geometry class, our proofs should be written in complete sentences.  You should break sections of a proof into paragraphs and use proper grammar.  There are some pedantic conventions for doing this that will be pointed out along the way.  Initially, this will be an issue that you may struggle with, but you will get the hang of it.
\item There will be many situations where you will want to refer to an earlier definition, problem, theorem, corollary, or lemma.  In this case, you should reference the statement by number, but it is also helpful to the reader to summarize the statement you are citing.  For example, you might write something like, ``By Theorem~\ref{thm:two consecutive ints}, the sum of two consecutive integers is odd, and so\ldots" or ``By the definition of divides (Definition~\ref{def:divides}), it follows that\ldots". One thing worth pointing out is that if we are citing a definition, theorem, or problem by number, we should capitalize Definition, Theorem, or Problem, respectively (e.g., see examples above). Otherwise, we do not capitalize these words (e.g., ``By the definition of divides\ldots").
\item There will be times when we will need to do some basic algebraic manipulations.  You should feel free to do this whenever the need arises.  But you should show sufficient work along the way.  You do not need to write down justifications for basic algebraic manipulations (e.g., adding 1 to both sides of an equation, adding and subtracting the same amount on the same side of an equation, adding like terms, factoring, basic simplification, etc.).  
\item On the other hand, you do need to make explicit justification of the logical steps in a proof.  As stated above, you should cite a previous definition, theorem, etc. when necessary.
\item Similar to making it clear where your proof begins, you should indicate where it ends.  It is common to conclude a proof with the standard ``proof box" ($\square$ or $\blacksquare$).  This little square at end of a proof is sometimes called a \textbf{tombstone} or \textbf{Halmos symbol} after Hungarian-born American mathematician \href{https://en.wikipedia.org/wiki/Paul_Halmos}{Paul Halmos} (1916--2006).
\end{itemize}

It is of utmost importance that you work to understand every proof.  Questions---asked to your instructor, your peers, and yourself---are often your best tool for determining whether you understand a proof.  Another way to help you process and understand a proof is to try and make observations and connections between different ideas, proof statements and methods, and to compare various approaches. Observations might sound like some of the following:
\begin{itemize}
\item When I tried this proof, I thought I needed to \ldots But I didn't need that, because \ldots
\item I think that \ldots is important to this proof, because \ldots
\item When I read the statement of this theorem, it seemed similar to this earlier theorem. Now I see that it [is/isn't] because \ldots
\end{itemize}

If you would like additional guidance before you dig in, look over the guidelines in Appendix~\ref{appendix:elements_of_style}: Elements of Style for Proofs. It is suggested that you review this appendix occasionally as you progress through the book as some guidelines may not initially make sense or seem relevant. 

Now, go have fun and start exploring mathematics!

\epigraph{Our greatest glory is not in never falling, but in rising every time we fall.}{Confucius, philosopher}

\end{section}